Gravitational Lens Recovery with \Glass: How to measure the mass profile and shape of a lens

### 2013-07-29 limits/ms.tex
editing lensing-theory part

  cluster lenses\citep{2009ApJ...690..154S,2013arXiv1304.2393S}, but

@@

\section{Theoretical Background}\label{sec:theory} %-----------------------------------------------------

\subsection{Lensing basics}\label{sec:lensing_basic} %-----------------------------------------------------

The lens equation:
%
\begin{equation}
\vec\beta = \vec\theta - \frac{D_{LS}}{D_S}\vec\alpha(\vec\theta)
\label{eqn:lens_equation}
\end{equation}
%
maps an observed image position $\vec\theta$ to a source position $\vec\beta$
via the correction factor \citep[e.g.][]{1992grle.book.....S}:
%
\begin{equation}
\vec\alpha(\vec\theta) = \frac{4GD_L}{c^2} \int \Sigma(\vec\theta')\frac{(\vec\theta - \vec\theta')}{\ |\vec\theta - \vec\theta'|^2}d^2\vec\theta'
\end{equation}
%
which accounts for the gravitational lensing effect of an intervening mass
(where $G$ is Newton's gravitational constant, and $c$ is the speed of light).
For our purposes we make use of the thin mass sheet approximation and assume
that the lensing mass is infinitely thin compared with the distance between the
source and the lens $D_{LS}$ and the lens and the observer $D_{L}$. These are
angular diameter distances where $D_* = (c/H_0)d_*$, $H_0$ is the Hubble constant, and $d_*$ is a
dimensionless number that depends on cosmology.  The lens can then be thought
of as a projected surface density $\Sigma$ which diverts the path of a photon
instantaneously through the bending angle $\vec\alpha$.

Since we will later want to model the density distribution with a computer it
will be convenient to choose units that make the relevant quantities of order
unity.  We therefore measure lengths in light years, time in years, positions
in arcseconds, and choose $c=1$ and $4\pi G = N^2$, where $N^2 \equiv 206,265$
arcsec/rad. The mass unit is then $11.988\ \Msun$. It will also be useful to
define a proxy to the Hubble constant $\zeta \equiv N^2 H_0$.

One way to understand the lens equation is via Fermat's principle. We can think of light as travelling only 
along extremum paths where lensed images occur \citep{1986ApJ...310..568B}. Such paths occur at the extrema of the photon {\it arrival time surface} $t(\vec\theta)$ that depends on the geometric path the photon takes and the general relativistic gravitational time dilation
due to a thin lens at redshift $z_L$. In its complete form, with all units
and cosmology included, we have \citep{1986ApJ...310..568B}:
%
\begin{eqnarray}
N^2ct(\vec\theta) & = & (1+z_L)\frac{D_{L}D_{S}}{D_{LS}}\frac12 |\vec\theta - \vec\beta|^2 \nonumber \\
& & - (1+z_L)\frac{4GD_{L}^2}{c^2}\int \Sigma(\vec\theta') \ln |\vec\theta-\vec\theta'| d^2\vec\theta'
\label{eqn:full_arrival_time}
\end{eqnarray}
%
where the factor of $D_L^2$ in the second term comes from the fact that $\Sigma$
has units of \Msun/lyr$^2$. We can clean this up by writing down a dimensionless time
delay:
%
\begin{equation}
\tau = \left[ (1+z_L) d_L\right]^{-1}\zeta t
\label{tau}
\end{equation}
%
in terms of our previous definitions. If we
further define a dimensionless density:
%
\begin{equation}
\kappa_\infty = \frac{4\pi G}{c^2}\frac{c}{H_0}d_L\Sigma
              = \frac{d_L}{\zeta}\Sigma
\end{equation}
%
and a lensing potential:
%
\begin{equation}
\psi(\vec\theta) = \frac1\pi \int \kappa_\infty(\vec\theta') \ln|\vec\theta - \vec\theta'| d^2\vec\theta'\
\label{lensing potential}
\end{equation}
%
we can express \eqnref{eqn:full_arrival_time} very compactly as:
%
\begin{equation}
\tau(\vec\theta) = \frac12 \xi |\vec\theta-\vec\beta|^2 - \psi(\vec\theta)
\label{arrival time}
\end{equation}
%
where $\xi \equiv D_{S}/D_{LS}$. We explicitly write $\kappa_\infty$ to remind ourselves
that there is no source distance factor involved. This will be useful later when we consider
multiple sources. Notice that the constraint $\nabla \tau = 0$ -- i.e. extrema in the arrival time surface -- recovers the lens equation (\eqnref{eqn:lens_equation}).

@@

>>>>

  cluster lenses \citep{2009ApJ...690..154S,2013arXiv1304.2393S}, but

@@

\section{Formulation}\label{sec:theory}

\subsection{Lensing essentials}\label{sec:lensing_basic}

We follow \cite{1986ApJ...310..568B} with minor differences in notation
notation.

The lens equation
\begin{equation}
\begin{aligned}
\vec\beta &= \vec\theta - \frac{D_{LS}}{D_S}\vec\alpha(\vec\theta) \\
\vec\alpha(\vec\theta) &= \frac{4G}{c^2D_L} \int \Sigma(\vec\theta')
                          \frac{(\vec\theta - \vec\theta')}
                          {\ |\vec\theta - \vec\theta'|^2} \, d^2\vec\theta'
\end{aligned}
\label{eqn:lens_equation}
\end{equation}
maps an observed image position $\vec\theta$ to a source position $\vec\beta$.
The $D$ factors are angular diameter distances, as in the previous section.
The lens can then be thought of as a projected surface density
$\Sigma$ which diverts the path of a photon instantaneously through
the bending angle $\vec\alpha$.

which accounts for the gravitational lensing effect of an intervening mass
(where $G$ is Newton's gravitational constant, and $c$ is the speed of light).
For our purposes we make use of the thin mass sheet approximation and assume
that the lensing mass is infinitely thin compared with the distance between the
source and the lens $D_{LS}$ and the lens and the observer $D_{L}$. These are
angular diameter distances where $D_* = (c/H_0)d_*$, $H_0$ is the Hubble constant, and $d_*$ is a
dimensionless number that depends on cosmology.

\begin{equation}
D_{LS} = \frac c{H_0} \frac{1+z_S}{1+z_L} \int_{z_L}^{z_S}
                      \frac{dz}{\sqrt{\Omega_m(1+z)^3 + \Omega_\Lambda}}
\end{equation}

One way to understand the lens equation is via Fermat's principle. We
can think of light as travelling only along extremum paths where
lensed images occur. Such paths occur at
the extrema of the photon {\it arrival time\/} $t(\vec\theta)$
that depends on the geometric path the photon takes and the general
relativistic gravitational time dilation due to a thin lens at
redshift $z_L$.
\begin{equation}
\begin{aligned}
\frac{ct(\vec\theta)}{(1+z_L)D_{L}}
&= {\textstyle\frac12} |\vec\theta - \vec\beta|^2
   \times \frac{D_{S}}{D_{LS}} \\
&- \frac{4GD_L}{c^2}
   \int \Sigma(\vec\theta') \ln |\vec\theta-\vec\theta'| \, d^2\vec\theta'
\label{full arrival time}
\end{aligned}
\end{equation}

\begin{equation}
\kappa(\vec\theta) \equiv \frac{4\pi GD_L}{c^2} \Sigma (\vec\theta)
\end{equation}

\begin{equation}
\kappa \sim \frac{z_L\,\Sigma}{10\rm\,kg\,m^{-2}}
\end{equation}

\begin{equation}
\tau(\vec\theta) = {\textstyle\frac12} |\vec\theta - \vec\beta|^2
                   \times \frac{D_{S}}{D_{LS}}
                 - \frac1\pi \int \kappa (\vec\theta')
                   \ln|\vec\theta - \vec\theta'| d^2\vec\theta'\
\label{arrival time}
\end{equation}
The scaled arrival time $\tau$ is like a solid angle.  It is of order
the area (in steradians) of the full lensing system --- the expression
$\sim|\vec\theta - \vec\beta|^2$ is of order the image-separation
squared, and the other terms are of similar size.  For this reason, is
convenient to measure $\tau$ in arcsec$^{2}$.

, with all units
and cosmology included, we have 
%
\begin{eqnarray}
N^2ct(\vec\theta) & = & (1+z_L)\frac{D_{L}D_{S}}{D_{LS}}\frac12 |\vec\theta - \vec\beta|^2 \nonumber \\
& & - (1+z_L)\frac{4GD_{L}^2}{c^2}\int \Sigma(\vec\theta') \ln |\vec\theta-\vec\theta'| d^2\vec\theta'
\label{eqn:full_arrival_time}
\end{eqnarray}
%
where the factor of $D_L^2$ in the second term comes from the fact that $\Sigma$
has units of \Msun/lyr$^2$. We can clean this up by writing down a dimensionless time
delay:
%
\begin{equation}
\tau = \left[ (1+z_L) d_L\right]^{-1}\zeta t
\label{tau}
\end{equation}
%
in terms of our previous definitions. If we
further define a dimensionless density:
%
\begin{equation}
\kappa_\infty = \frac{4\pi G}{c^2}\frac{c}{H_0}d_L\Sigma
              = \frac{d_L}{\zeta}\Sigma
\end{equation}
%
and a lensing potential:
%
\begin{equation}
\psi(\vec\theta) = 
\label{lensing potential}
\end{equation}
%
we can express \eqnref{eqn:full_arrival_time} very compactly as:
%

%
where $\xi \equiv D_{S}/D_{LS}$. We explicitly write $\kappa_\infty$ to remind ourselves
that there is no source distance factor involved. This will be useful later when we consider
multiple sources. Notice that the constraint $\nabla \tau = 0$ -- i.e. extrema in the arrival time surface -- recovers the lens equation (\eqnref{eqn:lens_equation}).

@@

Degeneracy first formally noted by \cite{1985ApJ...289L...1F}.  Most
important is mass sheet.  But in fact, there are infinitely many
\citep{2000AJ....120.1654S}.
Also shape degeneracies \citep{2006ApJ...653..936S,2013arXiv1306.4675S}
Suppressed by redshift contrast
\citep{1998AJ....116.1541A,2009ApJ...690..154S}
but not completely \citep{2008MNRAS.386..307L}.
